---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![Build Status](https://travis-ci.org/systemincloud/rly.svg?branch=master)](https://travis-ci.org/systemincloud/rly)
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/rly)](https://cran.r-project.org/package=rly)
[![](https://cranlogs.r-pkg.org/badges/rly)](https://cran.r-project.org/package=rly)
[![codecov](https://codecov.io/gh/systemincloud/rly/branch/master/graph/badge.svg)](https://codecov.io/gh/systemincloud/rly)
[![PayPal donation](https://img.shields.io/badge/donation-PayPal-red.svg)](https://www.paypal.com/cgi-bin/webscr?cmd=_donations&business=UR288FRQUSYQE&item_name=rly+-+R+Lex+and+Yacc&currency_code=USD&source=url)

# rly

Tools to Create Formal Language Parsers

## Introduction

{rly} is a 100% R implementation of the common parsing tools [`lex` and `yacc`](http://dinosaur.compilertools.net/).

`lex` is a "lexical analyzer generator". It's core functionality is to split up an input stream into more usable elements. You can think of it as a tools to help identify the interesting components in a text file, such as `->` or `%>%` in an R script.

`yacc` is "yet another compiler-compiler" and the main task for it is to take the tokens `lex` provides and process them contextually.

Together, you can use them to:

- define what tokens a given language/input stream will accept
- define what R code should be executed as a language file/input stream (e.g. a program) is parsed.

This project is a R clone of [Ply](https://github.com/dabeaz/ply).

## Installation

```{r eval=FALSE}
devtools::install_github("systemincloud/rly")
```

## Usage

{rly} consists of two files : `lex.R` and `yacc.R`. There's quite a bit "going on" in those two files and they may be helpful examples of how to structure {R6} classes.

```{r eval=TRUE}
library(rly)
```

## Examples

The `demo` directory contains several different examples. Here are some more.

### Finding Tokens

We can build a "lexer" to pull out URLs from text input. Sure, we could just use `stringi::stri_match_all_regex()` for this particular example, but the intent is to work with a known, straightforward domain to see how "lexing" works:

```{r url-lex-01}
# Define what tokens we accept. In this case, just URLs

TOKENS <- c("URL")

# Build our "lexing" rules. This is an {R6} class. 
# If you've never played with {R6} classes, head on
# over to <https://cran.rstudio.com/web/packages/R6/>
# to learn more about it and also take a look at the
# packages that depend on/use it (which may help you
# grok {R6} a bit better.)

URLexer <- R6::R6Class(
  
  classname = "Lexer",
  
  public = list(
    
    # tell it abt the tokens we accept
    
    tokens = TOKENS, 
    
    # we use the t_ prefix to identify that this is the 
    # "matcher" for the token and then give it the regular
    # expression that goes with this token. The URL
    # regex is a toy one that says to match http or https
    # strings until it hits a space.
    # 
    # The `t` parameter is the the full context of the token
    # parser at the time it gets to this token.
    #
    # here, we're just printing a message out and continuing
    # but we could do anything we (programmatically) want
    
    t_URL = function(re = 'http[s]*://[^[:space:]]+', t) {
      message("Found URL: ", t$value) # Be verbose when we find a URL
      return(t) # we need to return the potentially modified token
    },
    
    # whenever a newline is encounterd we increment a line #
    # counter. this is useful when providing contextual errors
    
    t_newline = function(re='\\n+', t) {
      t$lexer$lineno <- t$lexer$lineno + nchar(t$value)
      return(NULL)
    },
    
    # the goal of the lexer is to give us valid input 
    # but we can ignore errors if we're just looking for 
    # certain things (like URLs)
    
    t_error = function(t) {
      t$lexer$skip(1)
      return(t)
    }
    
  )
)

# Create out lexer

lexer  <- rly::lex(URLexer)

# Feed it some data

lexer$input(s = "
http://google.com     https://rstudio.org/
Not a URL  https://rud.is/b Another non-URL
https://r-project.org/



https://one.more.url/with/some/extra/bits.html


")

# We'll put found URLs here (rly inefficient)
found_urls <- character(0) 

# keep track of the # of invalid token info (also inefficient)
invalid <- list()

# Now, we'll iterate through the tokens we were given

repeat {
  
  tok <- lexer$token() # get the next token
  
  if (is.null(tok)) break # no more tokens, done with lexing
  
  switch(
    tok$type,
    
    # Do this when we find a token identified as a `URL`
    URL = found_urls <- append(found_urls, tok$value),
    
    # Do this whenever we find an invalid token
    error = invalid <- append(invalid, list(data.frame(
      bad_thing = tok$value,
      stream_pos = tok$lexpos,
      line = tok$lineno,
      stringsAsFactors = FALSE
    )))
  )
  
}

invalid <- do.call(rbind.data.frame, invalid)

nrow(invalid) # number of errors

head(invalid, 10) # it'll be clear we never told it abt whitespace

found_urls # the good stuff
```

We can extend this to do things with different types of URIs (not necessarily "http"-ish URLs)

```{r url-lex-02}
# we'll define different token types for HTTP URLs, HTTPS URLs and
# MAILTO URLs

TOKENS <- c("HTTP_URL", "HTTPS_URL", "MAILTO_URL")

URLexer <- R6::R6Class(
  
  classname = "Lexer",
  
  public = list(
    
    tokens = TOKENS, 

    # three different token regexes
    
    t_HTTPS_URL = function(re = 'https://[^[:space:]]+', t) {
      message("Found HTTPS URL: ", t$value)
      return(t)
    },

    t_MAILTO_URL = function(re = 'mailto:[^[:space:]]+', t) {
      message("Found MAILTO URL: ", t$value)
      return(t)
    },

    t_HTTP_URL = function(re = 'http://[^[:space:]]+', t) {
      message("Found HTTP URL: ", t$value)
      return(t)
    },

    t_error = function(t) {
      t$lexer$skip(1) # if we don't do this the lexer will error out on tokens we don't match (which is usually what we want)
      return(t)
    }
    
  )
)

# Create out lexer

lexer  <- rly::lex(URLexer)

# Feed it some data

lexer$input(s = "
http://google.com     https://rstudio.org/
Not a URL  https://rud.is/b Another non-URL mailto:fred@example.com?subject=Hello 
https://r-project.org/

mailto:steve@example.com

https://one.more.url/with/some/extra/bits.html


")

http_urls <- character(0) 
https_urls <- character(0) 
mailto_urls <- character(0) 

repeat {
  
  tok <- lexer$token() # get the next token
  
  if (is.null(tok)) break # no more tokens, done with lexing
  
  switch(
    tok$type,
    
    HTTP_URL = http_urls <- append(http_urls, tok$value),
    HTTPS_URL = https_urls <- append(https_urls, tok$value),
    MAILTO_URL = mailto_urls <- append(mailto_urls, tok$value)
    
  )
  
}

http_urls

https_urls

mailto_urls
```

### Calculator Example

Here is an example showing a {rly} implementation of a calculator with variables.

```{r echo=FALSE}
Sys.setenv("__R_CHECK_LENGTH_1_CONDITION_"=FALSE)
Sys.setenv("__R_CHECK_LENGTH_1_LOGIC2_"=FALSE)

```

Let's bootstrap our tokenizer/lexer:

```{r eval=TRUE}
TOKENS = c('NAME', 'NUMBER')
LITERALS = c('=', '+', '-', '*', '/', '(', ')') # these are "LEXEMES" (ref: https://stackoverflow.com/questions/14954721/what-is-the-difference-between-a-token-and-a-lexeme)

Lexer <- R6::R6Class(
  
  classname = "Lexer",
  
  public = list(
    
    tokens = TOKENS,
    
    literals = LITERALS,
    
    t_NAME = '[a-zA-Z_][a-zA-Z0-9_]*',
    
    t_NUMBER = function(re='\\d+', t) {
      t$value <- strtoi(t$value)
      return(t)
    },
    
    t_ignore = " \t",
    
    t_newline = function(re='\\n+', t) {
      t$lexer$lineno <- t$lexer$lineno + nchar(t$value)
      return(NULL)
    },
    
    t_error = function(t) {
      cat(sprintf("Illegal character '%s'", t$value[1]))
      t$lexer$skip(1)
      return(t)
    }
    
  )
)
```

Now, we write our expression parser. Note that we use `TOKENS` and `LITERALS` from the lexer we just wrote so the parser has some context for what gets passed to it as the tokeneizer emits bits to it:

```{r eval=TRUE}
Parser <- R6::R6Class(
  
  classname = "Parser",
                      
  public = list(
    
    tokens = TOKENS,
    
    literals = LITERALS,
    
    # Parsing rules
    precedence = list(
      c('left', '+', '-'),
      c('left', '*', '/'),
      c('right', 'UMINUS')
    ),
    
    # dictionary of names (can be inefficient but it's cool here)
    names = new.env(hash=TRUE),
    
    # One type of "statement" is NAME=expression
    
    p_statement_assign = function(doc='statement : NAME "=" expression', p) {
      self$names[[as.character(p$get(2))]] <- p$get(4)
    },
    
    # Another type of "statement" is just an expression
    
    p_statement_expr = function(doc='statement : expression', p) {
      cat(p$get(2))
      cat('\n')
    },
    
    # Classic simple definition of an expression
    
    p_expression_binop = function(doc="expression : expression '+' expression
                                                  | expression '-' expression
                                                  | expression '*' expression
                                                  | expression '/' expression", p) {
           if(p$get(3) == '+') p$set(1, p$get(2) + p$get(4))
      else if(p$get(3) == '-') p$set(1, p$get(2) - p$get(4))
      else if(p$get(3) == '*') p$set(1, p$get(2) * p$get(4))
      else if(p$get(3) == '/') p$set(1, p$get(2) / p$get(4))
    },
    
    # unary minus is a special case we need to handle 
    # see https://www.ibm.com/support/knowledgecenter/en/SSLTBW_2.3.0/com.ibm.zos.v2r3.bpxa600/bpxa698.htm
    # for %prec explanation
    # note order does kinda matter in both lexer and parser rule specs
    
    p_expression_uminus = function(doc="expression : '-' expression %prec UMINUS", p) {
      p$set(1, -p$get(3))
    },
    
    # parnens expression
    
    p_expression_group = function(doc="expression : '(' expression ')'", p) {
      p$set(1, p$get(3))
    },
    
    p_expression_number = function(doc='expression : NUMBER', p) {
      p$set(1, p$get(2))
    },
    
    p_expression_name = function(doc='expression : NAME', p) {
      p$set(1, self$names[[as.character(p$get(2))]])
    },
    
    p_error = function(p) {
      if(is.null(p)) cat("Syntax error at EOF")
      else           cat(sprintf("Syntax error at '%s'", p$value))
    }
    
  )
)

lexer  <- lex(Lexer)
parser <- yacc(Parser)

# these will each end with `NULL` as that's how the `parser` signals it's done
parser$parse("3", lexer)

parser$parse("3 + 5", lexer)

parser$parse("3 + 5 * 10 - 100", lexer)

parser$parse("A + B * C - D", lexer) # valid lexical syntax but no data to work on; in a real calculator this wld error out

parser$parse("A + B * C - D = E", lexer) # invalid lexical syntax 

parser$parse("A = 1 + 2", lexer) # valid syntax, still no output b/c we just did assignment
parser$parse("A", lexer) 

invisible(parser$parse("B = 5", lexer)) # using invisible() only to suppress useless NULLs
invisible(parser$parse("C = 10", lexer))
invisible(parser$parse("D = 100", lexer))

parser$parse("A + B * C - D", lexer) 
```
